---
title: "Introduction to R - Session 3^[Please do not circulate without permission from the author.]"
author: "Matthias Kuppler^[Department of Social Sciences, University Siegen. Email: <matthias.kuppler@uni-siegen.de>.]"
date: "Version: `r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    df_print: paged
    number_sections: true
---

```{r setup, echo=FALSE}
# Set code chunk options
knitr::opts_chunk$set(eval = FALSE)
```

# Warm-up

**Part 1:** Please write code that creates the following vectors.

-   A numeric vector with values 15 to 8 in steps of 1: 15, 14, ..., 8

-   A numeric vector with values from 75 to 5 in steps of 5: 75, 70, ..., 5

-   A numeric vector of length 9 with values going from 24 to 20

-   A logical vector: TRUE TRUE FALSE TRUE TRUE

```{r warm-up-1}
15:8
seq(from = 75, to = 5, by = -5)
seq(from = 24, to = 20, length.out = 9)
rep(x = c(TRUE, TRUE, FALSE), length.out = 5)
```

**Part 2:** Please compute the mean and standard deviation of the vector my.num. Round the result to 2 digits.

```{r warm-up-2}
# Create my.num
my.num <- c(1:6, NA, 25:14, NA, seq(from = 5, to = 18, length.out = 8), NA)

# Compute mean and standard deviation
round(mean(x = my.num, na.rm = TRUE), digits = 2)
round(sd(x = my.num, na.rm = TRUE), digits = 2)
```

**Part 3:** Please use logical operators to test the following propositions...

-   The values of vector my.num1 are equal to 3

-   The vector my.num1 is longer than the vector my.num2

-   The mean and standard deviation of my.num1 are larger than the mean and standard deviation of my.num2

```{r warm-up-3}
# Set seed (so that everyone gets the same result)
set.seed(seed = 1234)

# Create vectors
my.num1 <- sample(x = 1:100, size = 50, replace = TRUE)
my.num2 <- sample(x = 1:100, size = 25, replace = TRUE)

# Test the propositions
my.num1 == 3
length(my.num1) > length(my.num2)
mean(my.num1, na.rm = TRUE) > mean(my.num2, na.rm = TRUE) &
  sd(my.num1, na.rm = TRUE) > sd(my.num2, na.rm = TRUE)

# To understand the results
mean(my.num1, na.rm = TRUE)
mean(my.num2, na.rm = TRUE)
sd(my.num1, na.rm = TRUE)
sd(my.num2, na.rm = TRUE)

# Extra: Number of elements in my.num1 == 3 (Why does this work?)
sum(my.num1 == 3)
```

# Task 1

**Part 1:** Please use ifelse() to set all uneven values in my.vec to 0.

```{r solution-1-1}
# Create vector
my.vec <- 2:18

# Set uneven values to 0
my.vec <- ifelse(my.vec %% 2 != 0, 0, my.vec)

# Inspect result
my.vec
```

**Part 2:** Please use an if-else statement to print "This vector keeps the doctor away!" if the vector contains the word "apple" and to print "This vector DOES NOT keep the doctor away!" if the vector does not contain the word "apple". (Tip: You can use *word %in% my.vec* to test whether word is in my.vec!)

```{r solution-1-2}
# Vectors to test
my.char1 <- c("banana", "potato", "apple", "peach")
my.char2 <- c("pineapple", "broccoli", "pear", "pomegranate")

# If-else statement
if("apple" %in% my.char1){
  print("This vector keeps the doctor away!")
} else{
  print("This vector DOES NOT keep the doctor away!")
}

if("apple" %in% my.char2){
  print("This vector keeps the doctor away!")
} else{
  print("This vector DOES NOT keep the doctor away!")
}
```

# Task 2

Please write an for() loop that calculates and prints the difference between value $i$ and value $i-1$ of vector my.num. The first output of the loop would be value 2 - value 1.

```{r solution-2}
# Vector to loop over
my.num <- 1:10

# Your loop
for(i in 2:length(my.num)){
  print(my.num[i] - my.num[i - 1])
}
```

# Task 3

Please write a function called *sd2* that takes a numeric vector *x* as its input and reports the (unbiased sample) standard deviation of *x* as its output. The standard deviation is a measure of variation around the mean. It is defined as

$$
SD = \sqrt{\frac{1}{N-1} \times \sum_{i=1}^{N} (x_i - \bar{x})^2}
$$

Where $N$ is the number of elements in vector *x* (i.e., its length), $\sum$ is the summation operator (i.e., the sum() function), $x_i$ are the values in vector *x*, and $\bar{x}$ is the mean of vector *x* (i.e., the mean() function). Finally, $\sqrt{}$ is the square root (i.e., the sqrt() function).

```{r solution-3}
# Create numeric vector
my.vec <- sample(x = 1:100, size = 50, replace = TRUE)

# Function
sd2 <- function(x, na.rm = TRUE){
  # Check whether input is numeric
  if(is.numeric(x) == FALSE){stop("Input vector not numeric.")}
  # Check na.rm option
  if(na.rm == TRUE){
    # Check whether x has missing values
    if(sum(is.na(x)) > 0){warning("Removed missing values of input vector.")}
    # Remove missing values
    x <- x[is.na(x) == FALSE]
  } else{
    # Issue warning
    if(sum(is.na(x)) > 0){warning("Input vector has missing values.")}
  }
  # Return sample standard deviation
  return(
    sqrt(sum((x - mean(x))^2) / (length(x) - 1))
  )
}

# Test function
sd2(x = my.vec)
sd(x = my.vec)  # Build-in R function
```

# Task 4

Please use the apply() function to...

-   Find the maximum value for each row of my.mat

-   Find the minimum value for each column of my.mat

```{r solution-4}
# Set seed (so that everyone gets the same results)
set.seed(seed = 1234)

# Create matrix
my.mat <- matrix(data = sample(x = 1:100, size = 16, replace = TRUE), nrow = 4)
my.mat

# Row maximum
apply(X = my.mat, MARGIN = 1, FUN = max)

# Column minimum
apply(X = my.mat, MARGIN = 2, FUN = min)
```
