---
title: "Introduction to R - Session 1 Solutions^[Please do not circulate without permission from the author.]"
author: "Matthias Kuppler^[Department of Social Sciences, University Siegen. Email: <matthias.kuppler@uni-siegen.de>.]"
date: "Version: `r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    df_print: paged
    number_sections: true
---

```{r setup, echo=FALSE}
# Set code chunk options
knitr::opts_chunk$set(eval = FALSE)
```

# Task 1

Please write the commands that create the following vectors. For some of these vectors, there are multiple ways to create them. If possible, avoid using the combine function c().

-   14 13 12 11

-   1 2 3 1 2 3

-   1 2 3 3 2 1

-   "a" "a" "b" "b" "c" "c"

-   "d" "e" "f" "g" "h"

-   3.2 3.4 3.6 3.8 4.0

-   FALSE TRUE FALSE TRUE FALSE

```{r solution-1}
# Vector 1
14:11

# Vector 2
rep(x = 1:3, times = 2)
rep(x = 1:3, length.out = 6)

# Vector 3
c(1:3, 3:1)

# Vector 4
rep(x = c("a", "b", "c"), each = 2)
rep(x = letters[1:3], each = 2)

# Vector 5
letters[4:8]

# Vector 6
seq(from = 3.2, to = 4, by = .2)
seq(from = 3.2, to = 4, length.out = 5)

# Vector 7
rep(x = c(FALSE, TRUE), length.out = 5)
```


# Task 2

Can you write a command that tests whether the values of a numeric vector are even? The command should return a logical vector of equal length as the numeric vector, where TRUE indicates that the value is even and FALSE indicates that the value is uneven. (Tip: You must use the modulus %%!).

```{r solution-2}
# Vector to test
my.vec <- 1:10

# Solution
my.vec %% 2 == 0
```


# Task 3

Please write the commands to accomplish the subsetting tasks described below.

```{r solution-3}
# Vectors to subset
my.num <- 21:34
my.char <- c("apple", "pear", "orange", "carrot", "peach", "cucumber", "banana", "carrot")

# Part 1: Select the 4th to 7th value of my.num
my.num[4:7]

# Part 2: Select the 3rd, 6th, and 9th value of my.num
my.num[c(3, 6, 9)]
my.num[seq(from = 3, to = 9, by = 3)]

# Part 3: Select all values equal to 21, 24, and 27
my.num[my.num %in% c(21, 24, 27)]
my.num[my.num %in% seq(from = 21, to = 27, by = 3)]

# Part 4: Select everything but the last value of my.num
my.num[-length(my.num)]

# Part 5: Select all uneven values of my.num
my.num[my.num %% 2 != 0]

# Part 6: Select only the vegetables in my.char (carrot, cucumber)
my.char[my.char == "carrot" | my.char == "cucumber"]

# Part 7: Select all values with an even index number
my.char[c(FALSE, TRUE)]
```

# Task 4

Please write a command that creates the following matrix.

|     |     |     |
|-----|-----|-----|
| "a" | "d" | "g" |
| "b" | "e" | "h" |
| "c" | "f" | "i" |

```{r solution-4}
matrix(data = letters[1:9], nrow = 3, ncol = 3, byrow = FALSE)
```


# Task 5

**Part 1:** Please create a data frame named "task.df" with two columns. The first column should be named "id" and should contain the integer values from 1 to 10. The second column should be named "random" and should contain 10 draws from a uniform distribution with minimum = 0 and maximum = 1. (Tip: Use the runif() function! Find help with ?runif)

The command set.seed() sets the seed for the random number generator so that we all will get the same results.

**Part 2:** Select all rows in which the column "random" is greater than 0.5.

```{r solution-5}
# Set seed
set.seed(seed = 1234)

# Part 1
task.df <- data.frame(id = 1:10, random = runif(n = 10, min = 0, max = 1))

# Part 2
task.df[task.df$random > .5, ]
```

# Task 6

Please try to execute the following subsetting commands on our list "my.list". You can use the [[]] and the \$ operator.

```{r solutions-6}
# Extract the data frame and assign it to an object called "df"
df <- my.list$my.df

# Extract the 3rd value of the vector
my.list[[2]][3]
my.list$my.vec[3]

# Extract the second element from the list within my.list
my.list[[5]][[2]]
my.list$my.lst$el2
```
