---
title: "Introduction to R - Session 3^[Please do not circulate without permission from the author.]"
author: "Matthias Kuppler^[Department of Social Sciences, University Siegen. Email: <matthias.kuppler@uni-siegen.de>.]"
date: "Version: `r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    df_print: paged
    number_sections: true
---

```{r setup, echo=FALSE}
# Set code chunk options
knitr::opts_chunk$set(eval = FALSE)
```

# Task 1

Please install the packages "haven" and "readxl" from CRAN. The "**haven**" package contains functions to import data from non-standard data file formats (e.g., STATA's .dta format). The "**readxl**" package allows you to import Excel's native .xlsx files. Please also load the packages after installing them.

```{r solution-1}
# Install package
install.packages("haven")
install.packages("readxl")

# Load packages
library(tidyverse)
library(haven)
library(readxl)
```

# Task 2

The code below creates a data frame with artificial data. Please use cut() and/or factor() to:

1.  Convert the column *sex* into a factor with the labels Male (if sex = 0) and Female (if sex = 1)

2.  Categorize the column *income* into 10%-percentiles

Please store the resulting new columns as new variables (*sex.cat*, *income.cat*). You can do this with my.df\$new.var \<- cut(...).

```{r solution-2}
# Create data frame (artificial data)
my.df <- data.frame(
  # IDs
  id = 1:1000,
  # Sex (categorical)
  sex = sample(x = 0:1, size = 1000, replace = TRUE),
  # Income (continuous) from lognormal distribution
  income = 1000 * rlnorm(n = 1000, meanlog = 0, sdlog = .5)
)

# Sex
my.df$sex.cat <- factor(
  x = my.df$sex,
  levels = 0:1,
  labels = c("Male", "Female")
  )

# Income
my.df$income.cat <- cut(
  x = my.df$income,
  breaks = quantile(
    x = my.df$income,
    probs = seq(from = 0, to = 1, by = .1)
    ),
  include.lowest = TRUE
)
```

# Loading the ALLBUS

```{r import-allbus}
# Load data
df <- read.table(file = "./pdata/ALLBUS-2021-Selection.txt", sep = "\t")
```


# Declaring missing values

```{r declare-missing}
# Define missing functions
## For all variables 
mis.decl <- function(x){
  ifelse(x < 0, NA, x)
}

## For income variables
mis.decl2 <- function(x){
  ifelse(x == -50, 0, x)
  ifelse(x < 0, NA, x)
}

# Apply function across columns of data set
df <- df %>% mutate(across(c(respid, age:pa01), ~ mis.decl(.))) # Exclude income variables
df <- df %>% mutate(across(di01a:hhincc, ~ mis.decl2(.))) # Only income variables
```

# Task 3
Please use the filter() verb to make the following selections of rows (participants) in the ALLBUS:

-   Select only participants who are in some form of employment (work = 1, 2, and 3), thereby excluding all participants who are unemployed (work = 4)
-   Select only participants with at least upper secondary education (iscd11 = 3 or higher)
-   Select only participants who are German citizens (german = 1)

```{r solution-3}
# Part 1
filter(.data = df, work != 4)
filter(.data = df, work %in% 1:3)

# Part 2
filter(.data = df, iscd11 >= 3)

# Part 3
filter(.data = df, german == 1)
```

# Task 4

Please select the following columns and store them in a new object called "df.small": respid, age, german, sex.

```{r solution-4}
df.small <- select(.data = df, respid, age, german, sex)
```

# Task 5

Please use mutate() to create the following new columns:

-   inc.log: The natural logarithm of the column di01a
-   pol.cat: A factor for the respondent's political leaning, derived from the column *pa01* via the following mapping

| pol.cat (factor levels) | pa01 (integer values) |
|-------------------------|-----------------------|
| Left                    | 1, 2                  |
| More left than right    | 3, 4, 5               |
| More right than left    | 6, 7, 8               |
| Right                   | 9, 10                 |

: Mapping pol to pol.cat

```{r solution-5}
df <- mutate(
  .data = df,
  # Log income
  inc.log = log(di01a),
  # Categorical political leaning
  pol.cat = cut(
    x = pa01,
    breaks = c(0, 2, 5, 8, 10),
    labels = c("Left", "More left than right", "More right than left", "Right")
    )
)
```

# The pipe %\>%

Please write code to perform the following sequence of data transformations and store the results in a new data frame called df.ted:

-   Select the following columns: respid, age, sex, german
-   Filter rows for which age \> 25
-   Create a new column with the normalized age (age.norm)
-   Compute the minimum and maximum of the normalized age

```{r solution-pipe}
df.ted <- select(.data = df, respid, age, sex, german)
df.ted <- filter(.data = df.ted, age > 25)
df.ted <- mutate(.data = df.ted, age.norm = age - mean(age, na.rm = TRUE))
df.ted <- summarize(
  .data = df.ted,
  min.age = min(age.norm, na.rm = TRUE),
  max.age = max(age.norm, na.rm = TRUE)
  )
```

# Task 6

Please use group_by to calculate the median of income (di01a) by

-   employment status (work)
-   political leaning (pol.cat)

```{r solution-6}
# Income by employment status
df %>% 
  group_by(work) %>% 
  summarize(inc.med = median(di01a, na.rm = TRUE))

# Income by political leaning
df %>% 
  group_by(pol.cat) %>% 
  summarize(inc.med = median(di01a, na.rm = TRUE))
```
